---
title: "Dealing with model non-identifiability issues when sampling"
output:
  workflowr::wflow_html:
    toc: false
date: "`r format(Sys.time(), '%d %B, %Y')`"
header_includes:
    - \usepackage{amsmath}
editor_options:
  chunk_output_type: console
---

```{r include=FALSE}
library(sparsefactor)
library(tictoc)
knitr::opts_chunk$set(eval = TRUE)
```

We first simulate gene expression data across 40 genes for 12 individuals. This data is explained by 2 sparse factors ($\pi_1=\pi_2=0.2$) and 2 dense factors ($\pi_3=\pi_4=0.8$). A sparse factor regulates a gene with probability 0.2, whereas a dense factor regulates a gene with probability 0.8. The gene-specific precision of the Gaussian noise is simulated from $\text{Gamma}{(1,0.1)}$. For details of the model, see Section 1 of the [derivation](assets/derivation.pdf).

```{r}
set.seed(0)
K <- 4 # number of factors
N <- 12 # number of individuals
G <- 40 # number of genes
data <- simulate.data(K=K, N=N, G=G,
                      pivec=c(rep(0.2, 2), rep(0.8, 2)),
                      alphavec=rep(1, 4),
                      taushape=1, taurate=0.1)
```

A Gibbs sampler is used to simulate the posterior. 50,000 samples are simulated, the first 10,000 of which are discarded as a burn-in.

```{r, echo=-c(3,5)}
N.TOT <- 50000
N.BURN <- 10000
tic()
samples <- burn.thin(gibbs(N.TOT, data$ymat,
                            pivec=c(rep(0.2, 2), rep(0.8, 2)),
                            ptaushape=1, ptaurate=1,
                            palphashape=1, palpharate=1, seed=0),
                      burn=N.BURN, thin=1)
toc()
```

To assess whether the sampler has converged to a stationary distribution, we look at trace plots of the simulated samples. As an example, here are the trace plots for the factor activations for individual 2:

```{r, echo=-1, fig.height=6, fig.width=9}
par(mfrow=c(2,2), mar=c(4,4,1,1.5))
j <- 2 # individual 2
for(k in 1:K) {
    vec = samples$fmat[,k,j]
    plot(0, 0, xlab="iteration", type="n", ylab=paste0("fmat[,",k,",",j,"]"),
         xlim=c(N.BURN + 1, N.TOT), ylim=c(min(vec), max(vec)))
    lines((N.BURN+1):N.TOT, vec, type="l")
}
```

For both factors 1 and 2, the simulated values appear to be a mixture of: (i) values sampled around a local mode around 2, (ii) values sampled around a local mode around -2, and (iii) values sampled around a local mode near 0, with comparatively larger variance.

The behaviour occurs due to the *non-identifiability* of the model. The gene expression level of gene $i$ of individual $j$ is modelled as
\[y_{ij} = \sum_{k=1}^K l_{ik}f_{kj} + e_{ij},\]
where $l_{ik}$ is the regulation strength of factor $k$ on gene $i$, $f_{kj}$ is the activation weight of factor $k$ for individual $j$, and \(e_{ij}\) is a noise term. For a fixed $k$, note that flipping the signs of $l_{ik}$ and $f_{kj}$ (for all $i$ and $j$) does not change the model. This explains the presence of two equivalent modes around 2 and -2 in the first two trace plots. This is an example of *sign switching*. If we take the mean across all samples, the estimated mean will be shrunk towards 0 due to the presence of the two modes. Thus, to obtain more accurate posterior summaries, it is desirable for only one of these modes to be present.

We can therefore deduce that $f_{1,2}$ and $f_{2,2}$ should have means near 0 and 2 (or equivalent, -2), in either order. However, in the trace plots, factor 1 and factor 2 both have samples simulated around both modes. This occurs because the roles of factor 1 and factor 2 are swapped at certain iterations of sampling. This is an example of *label-switching*: for fixed $k$ and $k'$, if $l_{ik}$ is swapped with $l_{ik'}$ and $f_{kj}$ is swapped with $f_{k'j}$ (for all $i$ and all $j$), the model does not change. This happens for $k=1,\,k'=2$, and also for $k=3,\,k'=4$.

To undo label and sign switching, a relabelling algorithm is used. The intuition is that each sample is relabelled (i.e. signs and labels are switched) such that the sampled distribution of each entry resembles a normal distribution, hence having only one mode. For details of this algorithm, see section 2.2 of the derivation. After relabelling, the trace plots are as follows:

```{r, echo=-c(1,3,4), fig.height=6, fig.width=9}
tic()
relabelled <- relabel(samples)
toc()
par(mfrow=c(2,2), mar=c(4,4,1,1.5))
for(k in 1:K) {
    vec = relabelled$fmat[,k,j]
    plot(0, 0, xlab="iteration", type="n", ylab=paste0("fmat[,",k,",",j,"]"),
         xlim=c(N.BURN + 1, N.TOT), ylim=c(min(vec), max(vec)))
    lines((N.BURN+1):N.TOT, vec, type="l")
}
```

The switching behaviour seems to be mostly removed. This is confirmed by the fact that the posterior densities appear to be unimodal:

```{r, echo=-1, fig.height=6, fig.width=9}
par(mfrow=c(2,2), mar=c(4,4,1,1.5))
for(k in 1:K) {
    plot(density(relabelled$fmat[,k,j]), main="", ylab=paste0("fmat[,",k,",",j,"]"))
}
```

Unfortunately, this relabelling algorithm is not always successful. For example, some switching behaviour remains for the factor activations of individual 1 (especially factor 3):

```{r, echo=-1, fig.height=6, fig.width=9}
par(mfrow=c(2,2), mar=c(4,4,1,1.5))
j <- 1
for(k in 1:K) {
    vec = relabelled$fmat[,k,j]
    plot(0, 0, xlab="iteration", type="n", ylab=paste0("fmat[,",k,",",j,"]"),
         xlim=c(N.BURN + 1, N.TOT), ylim=c(min(vec), max(vec)))
    lines((N.BURN+1):N.TOT, vec, type="l")
}
```

Further investigation is needed to determine whether this is caused by a flawed implementation, or an inherent deficiency of the method.

